Spatial and temporal variability in economic availability of residual biomass

### 2021-10-31 analysis/manuscript.Rmd
get ready for new distances

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990-2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
```

@@

Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020.  Our
strengths

@@

We used a number of publicly available datasets to develop the datasets
used in our analysis.  Most of the data are freely available from the World
Wide Web (WWW). Much of the data processing required for this study was 
manipulation of spatial
data.  This was done using the QGIS software @QGISorg2021.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 Tg/a  corresponding to a small wood pellet
mill, 400 Tg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 Tg/a  corresponding to a large ethanol plant.

@@

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as
\begin{equation}
\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
\end{equation}
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct  cycle time:  the
time required to travel from the mill, to the collection site in the forest,
and back to the mill.  This provides the basis for our estimates of haul cost.
Calculating average haul speed using an arithmetic mean leads to a biased
estimate of cycle time.

@@

#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
hourly_truck_rate <- 175.90 #$/hr
#hourly_truck_rate <- 108.82 #$/hr
#solid_volume_factor <- 0.4 #from fortsv5
solid_volume_factor <- 0.3 #from fortsv5
adj_trailer_capacity <- trailer_capacity * solid_volume_factor
specific_gravity <- 0.38 #lodgepole pine
odt_trailer <- specific_gravity * adj_trailer_capacity
truck_rate <- hourly_truck_rate / odt_trailer

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

Our focus in this study is on the spatial and temporal variation in biomass
collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
(roads, supervision, overhead, maintenance, and compliance) were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
excluding transport, is
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .

@@

Timber Tracks produces a publication describing forestry equipment hourly rates
@TT2020.  The rate reported for an 8 axle tridem tractor with a
b-train trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
the tractor-trailer combination used to haul biomass would cost a 
similar amount.

@@

## Residue yield (Mg · ha^-1^ ) by township.

@@

\begin{equation}
d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
\end{equation}

@@

\begin{equation}
H(d_b) = \beta_0 d_b ^ {\beta_1}
\end{equation}

@@

\begin{equation}
r(d_b,h) = \frac{d(d_b,h)}{200}
\end{equation}

@@

\begin{equation}
a(d_b,h) = \pi r(d_b,h)^2
\end{equation}

\begin{equation}
v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
\end{equation}

@@

@@

harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol, colour="Estimated harvest")) +
  labs(x = "year") +
  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
  geom_line()  +
  geom_line(aes(y=total, colour="Actual harvest"))  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
  scale_x_continuous(limits=c(1990,2015))
# 
# head(harv4)

@@

@@

cyclePlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Tg )") +
  labs(y = "marginal cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
   geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,14))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.  Each curve represents one year.  The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
mcPlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Tg )") +
  labs(y = "marginal cost ($·Mg^-1^ )") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,250)) 
```

@@

bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($·Mg^-1^ )", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))

@@

avePlot <- ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($·Mg^-1^ )") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,200))

@@

ave <- aveave %>%
  gt() %>%
  tab_header(
    title = "summary stats"
    ) %>%
  fmt_number(
    columns = c(capacity,aveavecost,sdavecost,cv),
    suffixing = TRUE,
    n_sigfig = 3
  ) %>% 
  cols_label(location = "Location",
             capacity = "Capacity (Tg/a)",
             aveavecost = "Mean ($·Mg^-1^)", 
             sdavecost = "Std. Dev. ($·Mg^-1^)", 
             cv = "C.V.")
```

Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
for each of the 16 years in the study period, and for the 3 alternative plant 
locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
years of minimum, median, and maximum provincial harvest volume in the period. 
The inter-year variation in marginal costs increases as the cumulative amount
of residue collected increases.  We can also see that the relative position of
the years in terms of cost varies between the plant locations:  *e.g.* the year 
2005 is not as good for Hinton, as it was for Edson and Whitecourt.

@@

Figure 12 shows the average cost ($·Mg^-1^) for  each location, capacity, and
year.  Overall, the average costs in Edson and Whitecourt are quite similar.
Hinton is usually more expensive across years and capacities.  The differences
become clearer at a capacity of 800 Tg, where Whitecourt is the low cost location
for most of the years.

Table ?? presents summary statistics over the 16 year for the locations and 
capacities.  It is clear that average costs and the inter-annual variability of 
costs increase with increasing capacity.  It is clear that the variation is costs
for Hinton is greater than for Whitecourt and Edson, particularly for the 800
Tg/a capacities.

@@

Figure 14 compares the supply areas for 2005, 2009, and 1990 (high, median, and
low harvest years).  The area required to supply the mills increases by a lot
in low timber harvest years, particularly for the 800 Tg/a capacity.  Similar
maps could be produced for the other locations and all 16 years, but that is
unecessary here.

@@

<!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
<!-- 16 years of the study period.  The year 2009 had the median overall harvest  -->
<!-- level in Alberta over that period, and is coloured red in order to illustrate -->
<!-- the marginal cycle time for a median year. -->

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800 000 Mg/a  capacity, 
because they reach into the 
Whitecourt area.

@@

# Reference list

@@

Table:  Road speeds (km/h).

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| no road           |     5  |       5  |           5   |

@@

```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
knitr::include_graphics("../img/network_service.png")
```

@@

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
suppressWarnings(print(harv_comparison))
```

```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
knitr::include_graphics("../img/residue_per_ha.png")
```

```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
suppressWarnings(print(cyclePlot))
```

```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
suppressWarnings(print(mcPlot))
```

```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location.", out.width = '100%'}
suppressWarnings(print(bPlot))
```

```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
suppressWarnings(print(avePlot))
```

```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with demands of 200, 400, and 800 Tg.  The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
knitr::include_graphics("../img/w2009.png")
```

\newpage
\blandscape
```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
knitr::include_graphics("../img/wcontours.png")
```
\elandscape

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
header-includes:
- \usepackage[left]{lineno}
- \usepackage{setspace}
- \linenumbers
geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
```

\newpage
\doublespacing

@@

Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020a.  Our
strengths

@@

We used a number of publicly available datasets to develop the datasets
used in our analysis.  Most of the data are freely available on the World
Wide Web (WWW). Much of the data processing required for this study was 
manipulation of spatial
data.  This was done using the QGIS software @QGISorg2021.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 Gg/a  corresponding to a small wood pellet
mill, 400 Gg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 Gg/a  corresponding to a large ethanol plant.

@@

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as
$$
\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
$$
where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct  cycle time:  the
time required to travel from the mill, to the collection site in the forest,
and back to the mill.  This provides the basis for our estimates of haul cost.
Calculating average haul speed using an arithmetic mean leads to a biased
estimate of cycle time.

@@

#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
#hourly_truck_rate <- 175.90 #$/hr
hourly_truck_rate <- 157.31 #$/hr tandem tractor with tandem trailer
#hourly_truck_rate <- 108.82 #$/hr
#solid_volume_factor <- 0.4 #from fortsv5
solid_volume_factor <- 0.3 #from fortsv5
adj_trailer_capacity <- trailer_capacity * solid_volume_factor
specific_gravity <- 0.38 #lodgepole pine
odt_trailer <- specific_gravity * adj_trailer_capacity
truck_rate <- hourly_truck_rate / odt_trailer

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
#comminution_cost <- 25.60 # $/odt
comminution_cost <- 26.82 # $/odt (Blackburn2020a)
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

Our focus in this study is on the spatial and temporal variation in biomass
collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost (@Blackburn2020a)
was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
(roads, supervision, overhead, maintenance, and compliance) were
set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
excluding transport, is
`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .

@@

Timber Tracks produces a publication describing forestry equipment hourly rates
@TT2020.  The rate reported for 
#an 8 axle tridem tractor with a b-train trailer was
tandem tractor with a tandem trailer was
`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
the tractor-trailer combination used to haul biomass would cost a 
similar amount.

@@

## Residue yield (Mg·ha^-1^ ) by township.

@@

$$
d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
$$

@@

$$
H(d_b) = \beta_0 d_b ^ {\beta_1}
$$

@@

$$
r(d_b,h) = \frac{d(d_b,h)}{200}
$$

@@

$$
a(d_b,h) = \pi r(d_b,h)^2
$$

$$
v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
$$

@@

## The average tree!

FMA holders are now required by Alberta to project average piece size in the 
detailed forest management plans for their FMAAs.  We use these piece sizes
projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
per tree.
The mean piece size for needleaf species is 0.278 m^3^ and 0.264 m^3^ for broadleaf
species.
For needleaf and broadleaf species to calculate the average dbh and height
for the most common species in each township for the group using the
taper functions from Ung *et al.* @Ung2013, by searching for the dbh
that yields these piece sizes for the FMAA utilization standards using
binary search.  For FMUs not associated
with an FMAA, or for FMAAs with plans written before this requirement we
use the average piece size from the DFMPs.

@@

# harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol)) +
#   labs(x = "year") +
#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
#   geom_line()  +
#   geom_line(aes(y=total))  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
#   scale_x_continuous(limits=c(1990,2015)) + 
#   guides(fill=guide_legend(title="harvest volume"))

# harv_comparison <- ggplot(data=harv5, aes(x=year)) +
#   labs(x = "year") +
#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
#   geom_line(aes(y=harvest_vol))  +
#   geom_line(aes(y=total))  +
#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
#   scale_x_continuous(limits=c(1990,2015)) + 
#   guides(fill=guide_legend(title="harvest volume"))
# 
# head(harv4)

harv5a <- harv5 %>% 
  select(year,harvest_vol,total) %>% 
  rename(actual = total, estimated = harvest_vol) %>% 
  pivot_longer(!year,names_to = "Volume", values_to = "sum")

harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum,colour = Volume)) + 
  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
  geom_line(aes(y =  sum)) +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
  scale_x_continuous(limits=c(1990,2015)) +
  theme(legend.position = c(0.9,0.3))

@@

#save(cum.odmt,file="../data/cum.odmt.RData")

@@

cyclePlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Gg )") +
  labs(y = "marginal cycle time (h)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
   geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,14))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.  Each curve represents one year.  The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
mcPlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Gg )") +
  labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,250)) 
```

@@

bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = bquote('cost ('~"$"%.%Mg^-1~')'))+
  labs(x = "year") + 
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))

@@

avePlot <- ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = bquote('average cost ('~"$"%.%Mg^-1~')')) +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,200))

@@

ave <- aveave %>%
  gt() %>%
  tab_header(
    title = "summary stats"
    ) %>%
  fmt_number(
    columns = c(capacity,aveavecost,sdavecost,cv),
    suffixing = TRUE,
    n_sigfig = 3
  ) %>% 
  cols_label(location = "Location",
             capacity = "Capacity (Gg/a)",
             aveavecost = "Mean ($/Mg", 
             sdavecost = "Std. Dev. ($/Mg)", 
             cv = "C.V.")

Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
for each of the 26 years in the study period, and for the 3 alternative plant 
locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
years of minimum, median, and maximum provincial harvest volume in the period. 
The inter-year variation in marginal costs increases as the cumulative amount
of residue collected increases.  We can also see that the relative position of
the years in terms of cost varies between the plant locations:  *e.g.* the year 
2005 is not as good for Hinton, as it was for Edson and Whitecourt.

@@

Figure 12 shows the average cost ($·Mg^-1^) for  each location, capacity, and
year.  Overall, the average costs in Edson and Whitecourt are quite similar.
Hinton is usually more expensive across years and capacities.  The differences
become clearer at a capacity of 800 Gg, where Whitecourt is the low cost location
for most of the years.

Table ?? presents summary statistics over the 26 year for the locations and 
capacities.  It is clear that average costs and the inter-annual variability of 
costs increase with increasing capacity.  It is clear that the variation is costs
for Hinton is greater than for Whitecourt and Edson, particularly for the 800
Gg/a capacities.

@@

Figure 14 compares the supply areas for 2005, 2009, and 1990 (high, median, and
low harvest years).  The area required to supply the mills increases by a lot
in low timber harvest years, particularly for the 800 Gg/a capacity.  Similar
maps could be produced for the other locations and all 26 years, but that is
unecessary here.

@@

<!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
<!-- 26 years of the study period.  The year 2009 had the median overall harvest  -->
<!-- level in Alberta over that period, and is coloured red in order to illustrate -->
<!-- the marginal cycle time for a median year. -->

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800 Gg/a  capacity, 
because they reach into the 
Whitecourt area.

@@

# Reference list  {-}

@@

<!-- Table:  Road speeds (km/h). -->

<!-- | Road class      | loaded | unloaded | harmonic mean |  -->
<!-- |:----------------|-------:|---------:|--------------:| -->
<!-- | primary highways|    80  |       95 |           85  | -->
<!-- | other paved     |    65  |       80 |           70  | -->
<!-- | 2-lane gravel   |    50  |       70 |           60  |  -->
<!-- | 1-lane gravel or dirt  |    20  |       30 |           25  |  -->
<!-- | no road           |     5  |       5  |           5   | -->

Table:  Road speeds based on FPInnovations (km/h).

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    85  |       95 |           90  |
| other paved     |    77  |       86 |           81  |
| 2-lane gravel   |    60  |       67 |           63  | 
| 1-lane gravel or dirt  |    29  |       25 |           27  | 
| no road           |     5  |       5  |           5   |

@@

\newpage
\blandscape
```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.  The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.  The right panel indicates average harvestable volume (m³/ha) after non-merchantable stands have been filtered out.", out.width = '100%'}
knitr::include_graphics("../img/Merchantable Volume.png")
```
\elandscape

\newpage
\blandscape
```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
knitr::include_graphics("../img/network_service.png")
```
\elandscape

@@

\newpage
\blandscape
```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
suppressWarnings(print(harv_comparison))
```
\elandscape

```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
knitr::include_graphics("../img/residue.png")
```

\newpage
\blandscape
```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.  The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
suppressWarnings(print(cyclePlot))
```
\elandscape

\newpage
\blandscape
```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.  The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
suppressWarnings(print(mcPlot))
```
\elandscape

\newpage
\blandscape
```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015.", out.width = '100%'}
suppressWarnings(print(bPlot))
```
\elandscape

\newpage
\blandscape
```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
suppressWarnings(print(avePlot))
```
\elandscape

```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.  The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
knitr::include_graphics("../img/w2009.png")
```

\newpage
\blandscape
```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
knitr::include_graphics("../img/Cumulative Contours Whitecourt.png")
```
\elandscape
